The Spatial and Temporal Domains of Modern Ecology

### 2017-12-15 paper/main.tex
Edits to main text. Updated figure 1 and figure 1 code to include line indicated unreplicated space in panel D (duration)

\section*{Abstract}
\textbf{To properly understand ecological phenomena, it is necessary to observe them across a range of spatial and temporal scales. Ecologists first raised this point in the 1980s, and since then the ability to collect multi-scale observations has grown rapidly. To assess modern ecology's progress in addressing scale, we analyzed the resolution, extent, interval, and duration of observations in 348 observational studies published between 2004-2014. We found that the scale domains of observations are fairly narrow, and are collected primarily with conventional field techniques. In the spatial domain, most observations have resolutions of $\leq$1 m2 and extents of $\leq$10,000 ha. In the temporal domain, most observations were either unreplicated or of low frequency ($\geq$1 month interval), and were made over relatively short durations ($\leq$1 year). Compared to prior meta-analyses from the 1980s and early 2000s, observational durations and resolutions remain largely unchanged, but intervals have become finer and extents larger. Despite such gains, a large gulf exists between the scales at which phenomena are actually observed, and the scales those observations ostensibly represent, revealing unobserved portions of space and time not truly captured by replicates and raising concerns about observational comprehensiveness. Adding to these concerns, scales were not clearly reported in most studies, suggesting that it is a minor consideration. Journals can help mitigate this problem by implementing scale reporting standards, which can spur ecologists to more rapidly adopt new observational technologies, and thereby close key gaps in current observational domains.}

@@

Our analysis focused on two dimensions of spatial scale, resolution (grain) and extent, and two of temporal scale, interval and duration (Table 1, and see SI for full definitions). Resolution is the area of an individual spatial replicate within which a complete measurement (as opposed to a sub-sample) of the feature of interest was made. Extent is the area enclosed by the outer-most spatial replicates, or, if the system or habitat being sampled was distinct from its surrounding matrix (e.g. forest patches in grassland habitats), the summed area of sampled patches. Interval refers to the average time elapsed between individual temporal replicates. Duration measures the time elapsed between the first and last temporal replicates, or, in the case of temporally unreplicated observations, the estimated time spent collecting the observation. We also assessed observational scales within two additional dimensions, \emph{actual} extent (the summed area of spatial replicates) and \emph{actual} duration (the summed observational time of temporal replicates). We evaluated these additional dimensions to gain insight into how much the actual scales of observation (i.e. how much space and time is covered by the measurement) differ from the scales that the observations are intended to cover. This difference may contain important information about how effectively ecological observations characterize ecological phenomena. First, the odds that observations do not adequately represent the phenomenon of interest may increase (due to inappropriate sampling) as the gap between the actual and intended observational scales increases. Second (and related), a larger gap may increase the likelihood of collecting censored data (the values collected by observations are not fully known), as the frequent complexity, non-linearity, and non-randomness of ecosystems is such that phenomena may resolve themselves in the space or time between observations.

@@

%In contrast, cost is less likely to dictate the choice of observational scale in studies of natural systems (considering, for example, that many global-scale, multi-decadal, satellite-derived datasets are free \cite[e.g.][]{tucker_nasas_2004,friedl_modis_2010}), therefore an assessment confined to such studies may provide a less biased picture of ecology's scale domains.

%We calculated these from 379 discrete observations reported within a 134 paper subset of 348
%randomly selected articles (from 42,918 total). An additional 62 papers that were cited as the
%source of observations in the selected papers were also reviewed. We confined our analysis to observations of “natural” (7), or non-experimentally manipulated, phenomena. We made this choice because experiments are costly and therefore tend to be fine-scaled in both space and time (7, 9), which could unduly bias our analysis.
%
%We extracted these scale dimensions from 379 observations reported within 134 papers selected from the top 30 (based on impact factor) ecology-themed journals. The papers were a subset of 348 that we reviewed in total, which we selected randomly from the 42,918 studies published between 2004-2014 . An additional 62 papers that were cited as the source of observations in the selected papers were also reviewed. We confined our analysis to observations of ``natural'' \cite{tilman_ecological_1989}, or non-experimentally manipulated, phenomena. We made this choice because experiments are costly and therefore tend to be fine-scaled in both space and time \cite{tilman_ecological_1989,kareiva_spatial_1988}, which could unduly bias our analysis.

@@

\noindent \textbf{Results}\\
\noindent \textbf{\emph{Domains}}\\
In terms of resolution, the majority (67\%) of observations were collected in plots of $<$1 m$^2$ resolution, 24\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 9\% in plots of $\geq$1 ha (Fig. 1A). The extent of 19\% of observations was $<$10 ha, 23\% covered 10-1,000 ha, 12\% 1,000-10,000 ha, 19\% 10,000-100,000 ha, 12\% 100,000-1,000,000 ha, and 15\% $>$1,000,000 ha (Fig. 1B).

@@

\begin{figure}[!ht]
%\begin{wrapfigure}{c}{1\textwidth}
\includegraphics[width=1\textwidth]{figures/kde45.pdf}
\vspace{-0.15 cm}
\caption{Kernel density estimates of observational densities within the domains defined by A) interval and resolution (of temporally replicated observations only), B) duration and extent, C) resolution and extent, and D) interval and duration (of temporally replicated observations). Density estimates were applied to the log-transformed values of each observational dimension, and density estimates are rescaled to represent percentages.}
\label{afoto1}
\end{figure}

@@

% the bulk of ecological observations span one month to one decade in time and covered total time periods of one minute to $<$1 month and areas of $\leq$1 m$^2$ (Fig. 2B). The next greatest concentration were instantaneous ($\leq$1 second) observations of $<$1 m$^2$ extent, followed by a fainter concentration centered on 1 hour duration and between 1 m$^2$ and 1 ha in extent. Very few observations had spatial extents $>$1,000,000 ha, and because these were collected by satellites that make near-instantaneous measurements, they tended to have aggregate durations of $\leq$1 hour. In contrast, the longest periods of observation were provided by paleo-ecological or long-term weather stations, both of which produce continuous, point-scale measurements (e.g. sediment cores and weather instrumentation).

@@

%Additional sources of uncertainty: rules for extracting scales; diversity of study types and ecological methods.

%Most studies provided insufficient methodological detail to judge whether sampling schemes were suitable matched to and representative of the phenomena they were measuring. The observational scales were not clearly reported in many studies, such that we had to estimate spatial resolution and extent in 62\% and 68\% of cases, respectively, and temporal interval and duration in 40\% and 86\% of cases. Although the confidence intervals around our scale estimates (Fig. 1) suggests that our findings are robust to this level of uncertainty, it would have likely confounded efforts to estimate effective scales.

%This analysis did not necessarily quantify the scales \emph{represented} by ecological observations (SI). Observations that were non-continuous in time or space (e.g. point-based field measurements) may effectively represent larger scales than our estimates suggest, due to phenomenon-dependent factors such as autocorrelation and representativeness of the sampling scheme \cite{underwood_experiments_1997, palmer_scale_1994,cao_comparison_2002, legendre_spatial_1993,collins_method_2000-1}.

%Adding to our motivation to record actual observational scales was the fact that 
\noindent \textbf{Discussion}\\
\noindent \textbf{\emph{Insights into the scale domains of modern ecology}}\\
Our results suggest that the scale domains of modern ecological observations are fairly narrow, and are collected primarily with conventional field techniques. Spatially, the majority of observations have grains of $\leq$1 m$^2$ and extents of $\leq$10,000 ha (Fig 1A;B). In the temporal domains, most observations are either un-replicated snapshots, or of low frequency ($\geq$1 month interval; Fig. 1C) and relatively short duration ($\leq$1 year; Fig. 1D; 2D). Contrasting observational dimensions reveals that larger extents are associated with larger plot sizes (Fig. 2C), while longer durations are associated with longer intervals (Fig. 2D). The latter association presumably reflects a cost-imposed tradeoff between sampling frequency and temporal duration that is characteristic of traditional field-based observation. The same tradeoff is also responsible for the inverse relationship between resolution and interval that dominates that domain space (Fig. 2A). As a result of these tradeoffs, there are notable ``holes'' in the domains defined by high frequency (daily to sub-daily intervals) observations having 1) high to moderate resolutions ($\geq$1 m$^2$ up to 100 ha; Fig. 2A) and 2) decadal or longer durations (Fig. 2B).

%there are prominent holes in the scale domains of ecology, notably in the spaces defined by observations having 1) daily to sub-daily intervals and resolutions $\geq$1 m$^2$ up to 100 ha; 2) decadal or longer durations and extents $\geq$10,000,000 ha, such that few observations are collected at 1) high spatial and temporal resolution (1 m$^2$, 2) over large areas for long time periods.

@@

It is therefore important to examine whether the scales of the phenomena being observed are adequately captured by the design of replicates. Our methods suggest one possible procedure for assessing the \emph{scale representativeness} of replicates: 1) measure auto-correlation (spatial or serial) in the replicates, 2) add the autocorrelation length to the replicate area/duration, 3) calculate an autocorrelation-adjusted actual extent/duration, and 4) plot where it falls between actual extent/duration and extent/duration. The distance between the adjusted actual value and the ostensible value can provide a measure of how well the replicates represent the intended scale of observation. If the gap remains large, then alternative sampling methods may be used to close it. For example, remote sensing provides wall-to-wall spatial coverage of a study area, erasing the difference between actual extent and extent. Furthermore, the interval of high-resolution imaging (higher resolution is preferred in images as it allows individual features to be better discerned \cite{dark_modifiable_2007, hay_comparison_2003}) is now approaching daily to sub-daily scales \cite{drusch_sentinel-2:_2012, hand_startup_2015}, allowing improved representation of spatial and temporal dynamics. For phenomena that can't be measured from space, either because they are not visible or because they require continuous observation, new approaches for collecting \emph{in situ} or near-surface observations (e.g. low-cost wireless sensors \cite{wolf_gsm-based_2012, collins_new_2006, porter_wireless_2005}, citizen observers \cite{dickinson_current_2012}, and autonomous vehicles \cite{anderson_lightweight_2013}) can be used to increase the spatial and temporal coverage of observations.

@@

%In contrast, cost is less likely to dictate the choice of observational scale in studies of natural systems (considering, for example, that many global-scale, multi-decadal, satellite-derived datasets are free \cite[e.g.][]{tucker_nasas_2004,friedl_modis_2010}), therefore an assessment confined to such studies may provide a less biased picture of ecology's scale domains.

%These results indicate that for most ecological observations are collected with spatially discontinuous replicates that sample very small portions of the area in which they are distributed.
%
%Very little research is conducted at high spatial and temporal resolutions over large areas or for long time periods, indicating that, despite the well-established understanding of the importance of multi-scale assessments for understanding ecological patterns and processes \cite{levin_problem_1992,wiens_spatial_1989}, efforts focused on larger scales are still relatively sparse within the discipline \cite{levin_problem_1992,wiens_spatial_1989}. 
%
%a concern that has been previously noted due to ecology's geographical bias towards anthropogenically undisturbed and temperate ecosystems \cite{martin_mapping_2012}. 
%
%As with spatial data, finely resolved temporal data can be aggregated to facilitate multi-scale analyses (and many of the reviewed studies that used automated sensing aggregated the resulting observations before analyzing them), whereas longer-interval data (e.g. annual biomass accumulation) cannot be disaggregated into shorter interval measurements (e.g. weekly biomass accumulation) without interpolation, which turns data into modeled, rather than direct, observations.

%In the coming years, rapid technological advances should increase the concentration of ecological observations in currently under-represented domains. The growing numbers of high-resolution satellite sensors, together with new analytical platforms that provide free access to large volumes of pre-processed data and computational power \cite{googleearthengine}, will lower technical barriers that have so far prevented ecologists from adopting this observational technology  \cite{pettorelli_satellite_2014}. Similarly, the advent of unmanned systems offers the ability to measure ecological features at high spatial and temporal frequencies over large areas \cite{anderson_lightweight_2013}, which were scales that were previously impractical to access. The ever-falling cost of sensor technology and the ubiquity of cell phones also means that ecologists, together with a growing army of citizen-scientists, have the unprecedented ability to make spatially dense, high frequency observations over large areas \cite{wolf_gsm-based_2012,collins_new_2006,porter_wireless_2005,dickinson_current_2012}.  A greater attentiveness to scale in general, including more meticulous documentation of observed dimensions, may help to facilitate the spread of ecological research to sparsely studied scales, while improving transferability of knowledge within the discipline.

%\paragraph*{Displayed math.} 
\bibliography{/Users/lestes/Dropbox/publications/fullbib}
\bibliographystyle{naturemag}

\noindent \textbf{Acknowledgements} This work was supported by funds from the Princeton Environmental Institute Grand Challenges program and the NASA New Investigator Program (NNX15AC64G). Erle Ellis, Jason Chang, and Labeeb Ahmed of the GLOBE Project (http://globe.umbc.edu) were supported by the U.S. National Science Foundation (1125210).

% Following is a new environment, {scilastnote}, that's defined in the
% preamble and that allows authors to add a reference at the end of the
% list that's not signaled in the text; such references are used in
% *Science* for acknowledgments of funding, help, etc.

% For your review copy (i.e., the file you initially send in for
% evaluation), you can use the {figure} environment and the
% \includegraphics command to stream your figures into the text, placing
% all figures at the end.  For the final, revised manuscript for
% acceptance and production, however, PostScript or other graphics
% should not be streamed into your compliled file.  Instead, set
% captions as simple paragraphs (with a \noindent tag), setting them
% off from the rest of the text with a \clearpage as shown  below, and
% submit figures as separate files according to the Art Department's
% instructions.

%\begin{scilastnote}
%\item This work was supported by funds from the Princeton Environmental Institute Grand Challenges program and the NASA New Investigator Program (NNX15AC64G). Erle Ellis, Jason Chang, and Labeeb Ahmed of the GLOBE Project (http://globe.umbc.edu) were supported by the U.S. National Science Foundation (1125210).
%\end{scilastnote}

\clearpage

>>>>

\section*{Abstract}
\textbf{To properly understand ecological phenomena, it is necessary to observe them across a range of spatial and temporal scales. Ecologists first raised this point in the 1980s, and since then the ability to collect multi-scale observations has grown rapidly. To assess modern ecology's progress in addressing scale, we analyzed the resolution, extent, interval, and duration of observations in 348 observational studies published between 2004-2014. We found that the scale domains of observations are fairly narrow, and are collected primarily with conventional field techniques. In the spatial domain, most observations have resolutions of $\leq$1 m2 and extents of $\leq$10,000 ha. In the temporal domain, most observations were either unreplicated or of low frequency ($\geq$1 month interval), and were made over relatively short durations ($\leq$1 year). Compared to prior meta-analyses from the 1980s and early 2000s, observational durations and resolutions remain largely unchanged, but intervals have become finer and extents larger. Despite such gains, a large gulf exists between the scales at which phenomena are actually observed, and the scales those observations ostensibly represent, revealing portions of space and time not truly captured by replicates and raising concerns about observational comprehensiveness. Adding to these concerns, scales were not clearly reported in most studies, suggesting that it is a minor consideration. Journals can help mitigate this problem by implementing scale reporting standards, which can spur ecologists to more rapidly adopt new observational technologies, and thereby close key gaps in current observational domains.}

@@

Our analysis focused on two dimensions of spatial scale, resolution (grain) and extent, and two of temporal scale, interval and duration (Table 1, and see SI for full definitions). Resolution is the area of an individual spatial replicate within which a complete measurement (as opposed to a sub-sample) of the feature of interest was made. Extent is the area enclosed by the outer-most spatial replicates, or, if the system or habitat being sampled was distinct from its surrounding matrix (e.g. forest patches in grassland habitats), the summed area of sampled patches. Interval refers to the average time elapsed between individual temporal replicates. Duration measures the time elapsed between the first and last temporal replicates, or, in the case of temporally unreplicated observations, the estimated time spent collecting the observation. We also assessed observational scales within two additional dimensions, \emph{actual} extent (the summed area of spatial replicates) and \emph{actual} duration (the summed observational time of temporal replicates). We evaluated these additional dimensions to gain insight into how much the actual scales of observation (i.e. how much space and time is covered by the measurement) differ from the scales that the observations are explicitly or implicitly intended to represent. This difference may contain important information about how effectively ecological observations characterize ecological phenomena. First, an increasing gap between actual and intended observational scale inherently implies greater interpolation or extrapolation of observed measurements, raising the odds of over-leveraging data. Second (and relatedly), since natural systems are frequently complex, non-linear, and non-random \cite{levin_ecosystems_1998,pringle_spatial_2017,rietkerk_regular_2008}, a larger gap may increase the likelihood of encountering unanticipated data challenges such as censored data (\emph{sensu} \cite{efron_efficiency_1977}), as phenomena may resolve themselves in the space or time between observations.

@@

@@

\noindent \textbf{Results}\\
\noindent \textbf{\emph{Domains}}\\
In terms of resolution, the majority (67\%) of observations were collected in plots of $<$1 m$^2$ resolution, 24\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 9\% in plots of $\geq$1 ha (Fig. 1A). The extent of 19\% of observations was $<$10 ha, 23\% covered 10-1,000 ha, 12\% 1,000-10,000 ha, 19\% 10,000-100,000 ha, 12\% 100,000-1,000,000 ha, and 15\% $>$1,000,000 ha (Fig. 1B). Bar widths in (C-D) indicate differences in scale between x-axis labels.

@@

\begin{figure}[!ht]
%\begin{wrapfigure}{c}{1\textwidth}
\includegraphics[width=1\textwidth]{../vignettes/figures/fig2.png}
\vspace{-0.15 cm}
\caption{Kernel density estimates of observational densities within the domains defined by A) interval and resolution (of temporally replicated observations only), B) duration and extent, C) resolution and extent, and D) interval and duration (of temporally replicated observations). Density estimates were applied to the log-transformed values of each observational dimension, and density estimates are rescaled to represent percentages. Letters in the plots denote the median values of different observational methods (f=field observations; a = automated sensing; r = remote sensing; p = paleo-observations). The grey shaded areas represent physically impossible domains (intervals greater than duration and resolutions greater than extent).}
\label{afoto1}
\end{figure}

@@

@@

\noindent \textbf{Discussion}\\
\noindent \textbf{\emph{Insights into the scale domains of modern ecology}}\\
Our results suggest that the scale domains of modern ecological observations are fairly narrow, and are collected primarily with conventional field techniques. Spatially, the majority of observations have grains of $\leq$1 m$^2$ and extents of $\leq$10,000 ha (Fig 1A;B). In the temporal domains, most observations are either un-replicated snapshots, or of low frequency ($\geq$1 month interval; Fig. 1C) and relatively short duration ($\leq$1 year; Fig. 1D; 2D). Contrasting observational dimensions reveals that larger extents are associated with larger plot sizes (Fig. 2C), while longer durations are associated with longer intervals (Fig. 2D). The latter association presumably reflects a cost-imposed tradeoff between sampling frequency and temporal duration that is characteristic of traditional field-based observation. The same tradeoff is also responsible for the inverse relationship between resolution and interval that dominates that domain space (Fig. 2A). As a result of these tradeoffs, there are notable ``holes'' in the domains defined by high frequency (daily to sub-daily intervals) observations having 1) high to moderate resolutions ($\geq$1 m$^2$ up to 100 ha; Fig. 2A) and 2) decadal or longer durations (Fig. 2B).

@@

It is therefore important to examine whether the scales of the phenomena being observed are adequately captured by the design of replicates. Our methods suggest one possible procedure for assessing the \emph{scale representativeness} of replicates: 1) measure auto-correlation (spatial or serial) in the replicates, 2) add the autocorrelation length to the replicate area/duration, 3) calculate an autocorrelation-adjusted actual extent/duration, and 4) plot where it falls between actual extent/duration and extent/duration. The distance between the adjusted actual value and the ostensible value can provide a measure of how well the replicates represent the intended scale of observation. Though increasing spatial or temporal coverage may not always be the goal of a study (e.g., when spatial or temporal autocorrelation is a measure of interest), if the gap between actual and ostensible values remains large, then alternative sampling methods may be used to close it. For example, remote sensing provides wall-to-wall spatial coverage of a study area, erasing the difference between actual extent and extent. Furthermore, the interval of high-resolution imaging (higher resolution is preferred in images as it allows individual features to be better discerned \cite{dark_modifiable_2007, hay_comparison_2003}) is now approaching daily to sub-daily scales \cite{drusch_sentinel-2:_2012, hand_startup_2015}, allowing improved representation of spatial and temporal dynamics. For phenomena that can't be measured from space, either because they are not visible or because they require continuous observation, new approaches for collecting \emph{in situ} or near-surface observations (e.g. low-cost wireless sensors \cite{wolf_gsm-based_2012, collins_new_2006, porter_wireless_2005}, citizen observers \cite{dickinson_current_2012}, and autonomous vehicles \cite{anderson_lightweight_2013}) can be used to increase the spatial and temporal coverage of observations.

@@

\bibliography{/Users/lestes/Dropbox/publications/fullbib}
\bibliographystyle{naturemag}

\noindent \textbf{Acknowledgements} This work was supported by funds from the Princeton Environmental Institute Grand Challenges program and the NASA New Investigator Program (NNX15AC64G). Erle Ellis, Jason Chang, and Labeeb Ahmed of the GLOBE Project (http://globe.umbc.edu) were supported by the U.S. National Science Foundation (1125210).
\clearpage
